---
title: "Effect sizes for Bayesian models"
output: 
  github_document:
    toc: true
    fig_width: 10.08
    fig_height: 6
  rmarkdown::html_vignette:
    toc: true
    fig_width: 10.08
    fig_height: 6
tags: [r, effect size, rules of thumb, guidelines, conversion]
vignette: >
  \usepackage[utf8]{inputenc}
  %\VignetteIndexEntry{Converting between indices of effect size}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: bibliography.bib
---


```{r message=FALSE, warning=FALSE, include=FALSE}
library(knitr)
options(knitr.kable.NA = '')
knitr::opts_chunk$set(comment=">")
options(digits=2)
```


Obtaining indices of effect size for Bayesian model is currently an issue, as no test statistics are present to help us re-compute such indices. The two alternatives (both available in `effectsize`) is 1) to compute [**standardized parameters**](https://easystats.github.io/effectsize/articles/standardize_parameters.html) or 2) obtain effect sizes by test-statistic approximation. This second method is described here.

```{r, warning=FALSE, message=FALSE}
library(ppcor)

df <- iris[, 1:4]  # Remove the Species factor

ppcor::pcor(df)$estimate[2:4, 1]  # Select the rows of interest
```

The goal is to retrieve coefficients similar to the above **(partial) correlations** for this multiple regression model: `Sepal.Length ~ Sepal.Width + Petal.Length + Petal.Width`. This can easily be achieved for frequentist models by converting the *t* statistic into a correlation:


```{r, warning=FALSE, message=FALSE}
model <- lm(Sepal.Length ~ Sepal.Width + Petal.Length + Petal.Width, data = df) 

parameters <- model_parameters(model)[2:4,]
convert_t_to_r(parameters$t, parameters$df_residual)
```


Let's start by fitting the Bayesian regression:

```{r, warning=FALSE, message=FALSE, eval = FALSE}
library(rstanarm)

model <- stan_glm(Sepal.Length ~ Sepal.Width + Petal.Length + Petal.Width, data = df) 
```

```{r, warning=FALSE, message=FALSE, echo = FALSE}
library(rstanarm)
model <- stan_glm(Sepal.Length ~ Sepal.Width + Petal.Length + Petal.Width, data = df, refresh = 0, chains = 2) 
```


- **Step 1: Get degrees of freedom (*df*):**

The first step is to retrieve the degrees of freedom. For that, the most straightfoward way is to take the ones of the corresponding frequentist model.






```{r, warning=FALSE, message=FALSE}

```



## References