---
title: "Effect sizes for logistic models"
output: 
  github_document:
    toc: true
    fig_width: 10.08
    fig_height: 6
  rmarkdown::html_vignette:
    toc: true
    fig_width: 10.08
    fig_height: 6
tags: [r, effect size, logistic, regression, glm, binomial]
vignette: >
  \usepackage[utf8]{inputenc}
  %\VignetteIndexEntry{Effect sizes for logistic models}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: bibliography.bib
---


```{r message=FALSE, warning=FALSE, include=FALSE}
library(knitr)
options(knitr.kable.NA = '')
knitr::opts_chunk$set(comment=">")
options(digits=2)
```

# Data

Let's start off with a simple example. We will simulate 100 observations of a normally distributed variable `x` (mean = 0 and SD = 1) and a grouping variable `y`, made of *zeros* and *ones*. Importantly, the mean difference between these two groups (zeros *vs.* ones) is of 1.

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)

data <- bayestestR::simulate_difference(n=100, d=1, names=c("y", "x"))

summary(data)
```


# *t*-test

Now, as we are interested in the difference between these two groups, we can first investigate it using a *t*-test.

```{r message=FALSE, warning=FALSE}
data %>%  
  ggplot(aes(x=y, y=x, fill=y)) +
  geom_boxplot() +
  see::theme_modern()

ttest <- t.test(x ~ y, data=data, var.equal=TRUE) %>% 
  parameters::parameters()
ttest
```

As we can see, this confirms our simulation specifications, the difference is indeed of 1.

# Cohen's *d*

Let's compute now, a traditional Cohen's *d* using the `effectsize` package. While this *d* should be close to 1, it should theoretically be a tiny bit larger, because it takes into account the (pooled) SD of the whole variable `x` (across the groups), which because of the difference is a bit larger than 1.

```{r message=FALSE, warning=FALSE}
sd(data$x)
```

We can compute the Cohen's *d* as follows:

```{r message=FALSE, warning=FALSE}
effectsize::cohens_d(data$x, data$y)
```

As expected, it's pretty close to 1 times the SD of the sample. Interestingly, one can estimate the Cohen's *d* directly from the result of the *t*-test, using the *t* statistic. We can convert it to a *d* using the `effectsize` package:

```{r message=FALSE, warning=FALSE}
effectsize::t_to_d(ttest$t, df_error=ttest$df)
```

Fortunately, they are quite close.

# Logistic model

Another way of investigating these differences is through the lens of a logistic regression. The main difference is that here, the group variable `y` becomes the outcome and `x` the predictor. Let's fit such model and investigate the parameters: 


```{r message=FALSE, warning=FALSE}
model <- glm(y ~ x, data=data, family="binomial")

parameters::parameters(model)
```

How to interpret this output? The coefficients of a logistic model are expressed in log-odds, which is a metric of probability. Using the [**modelbased**](https://github.com/easystats/modelbased) package, one can easily visualize this model:


```{r message=FALSE, warning=FALSE}
model %>% 
  modelbased::estimate_link() %>% 
  ggplot(aes(x=x, y=Predicted)) +
  geom_ribbon(aes(ymin=CI_low, ymax=CI_high), alpha=0.2) +
  geom_line(color="red", size=1) + 
  see::theme_modern()
```

We can see that the probability of `y` being 1 (vs. 0) increases as `x` increases. This is another way of saying that there is a difference of `x` between the two groups of `y`. We can visualize all of our this together as follows:


```{r message=FALSE, warning=FALSE}
# Convert the 'y' variable to a numeric so that it's plotted correctly
data_plot <- data
data_plot$y <- as.numeric(data$y)-1

# Plot
ggplot() +
  geom_boxplot(data=data_plot, aes(x=y, y=x, fill=y, group=y)) +
  geom_jitter(data=data_plot, aes(x=y, y=x, fill=y, group=y), width=0.1) +
  geom_line(data=modelbased::estimate_link(model), aes(x=Predicted, y=x), color="red", size=1) +
  see::theme_lucid()
```

You can notice that the red predicted probability line passes through `x=0` when `y=0.5`. This means that when `x=0`, the probability of the two groups is equal: it is the "middle" of the difference between them.

<!-- # Odds ratios to Cohen's *d* -->

<!-- There is *in theory* a formula to convert these log-odds ratios (the unit of the coefficients in logistic models) to standardized differences (like Cohen's *d*). The formula is: $$d = logodds * (\sqrt{3} / \pi)$$ -->


<!-- Let's see: -->

<!-- ```{r message=FALSE, warning=FALSE} -->
<!-- logodds <- parameters::parameters(model)$Coefficient[2] -->

<!-- effectsize::logodds_to_d(logodds) -->
<!-- ``` -->

<!-- Hum, it seems a bit far from the value computed above! Let's investigate. -->

<!-- # Simulation -->

<!-- We will simulate different datasets with a varying *difference* and *sample size* (number of observations). For each of this dataset, we will compute the "true" Cohen's *d*, as well as the *d* obtained from the conversion of the logistic coefficient. Note that the simulation takes about 30 s to run. -->

<!-- ```{r message=FALSE, warning=FALSE} -->
<!-- df <- data.frame() -->
<!-- for(d in seq(0.1, 2, length.out=20)){ -->
<!--   for(n in exp(seq(log(20), log(150), length.out=20))){ -->
<!--     data <- bayestestR::simulate_difference(n=n, d=-d, names=c("y", "x")) -->
<!--     params <- parameters::parameters(glm(y ~ x, data=data, family="binomial"))[2, ] -->
<!--     params$d <- d -->
<!--     params$n <- n -->
<!--     params$d_cohen <- effectsize::cohens_d(data$x, data$y)$Cohens_d -->
<!--     df <- rbind(df, params) -->
<!--   } -->
<!-- } -->
<!-- df$logodds <- df$Coefficient -->
<!-- df$d_from_coef <- abs(effectsize::logodds_to_d(df$logodds)) -->
<!-- ``` -->

<!-- ```{r message=FALSE, warning=FALSE} -->
<!-- df %>% -->
<!--   ggplot(aes(x=d_cohen, y=d_from_coef, alpha=n)) + -->
<!--   see::geom_point2(size=2) + -->
<!--   geom_abline(slope=1) + -->
<!--   coord_fixed(ratio = 1, xlim = c(0, 3), ylim = c(0, 3)) + -->
<!--   see::theme_modern() -->
<!-- ``` -->


<!-- Oops, it seems like this formula doesn't work well, as it tends to underestimate the Cohen's *d* (and this tendency interacts with the sample size). -->

<!-- # Adjusted metric -->

<!-- Can we come up with a better conversion algorithm than `d = logodds * 0.55` (`0.55` being the result of `sqrt(3) / pi`)? Let's try. We will fit a model to predict the Cohen's *d* based on the logodds and the sample size, and plot the prediction. -->

<!-- ```{r message=FALSE, warning=FALSE} -->
<!-- model <- lm(d_cohen ~ logodds * n, data=df) -->

<!-- df$d_from_coef_adjusted1 <- modelbased::estimate_response(model)$Predicted    -->

<!-- ggplot(df, aes(x=d_cohen, y=d_from_coef_adjusted1, alpha=n)) + -->
<!--   see::geom_point2(size=2) + -->
<!--   geom_abline(slope=1) + -->
<!--   coord_fixed(ratio = 1, xlim = c(0, 3), ylim = c(0, 3)) + -->
<!--   see::theme_modern() -->
<!-- ``` -->

<!-- Seems better already! But if we look at the residuals, we can see that there is still a non-linear relationship with the sample size: -->

<!-- ```{r message=FALSE, warning=FALSE} -->
<!-- df$resid <- residuals(model) -->

<!-- ggplot(df, aes(x=n, y=resid, alpha=d_from_coef_adjusted1)) + -->
<!--   see::geom_point2(size=2)  + -->
<!--   geom_hline(yintercept = 0) + -->
<!--   see::theme_modern() -->
<!-- ``` -->


<!-- Can we further try to eliminate this effect? -->

<!-- ```{r message=FALSE, warning=FALSE} -->
<!-- model <- lm(d_cohen ~ logodds * n * log(n), data=df) -->
<!-- df$d_from_coef_adjusted2 <- modelbased::estimate_response(model)$Predicted    -->

<!-- ggplot(df, aes(x=d_cohen, y=abs(d_from_coef_adjusted2), alpha=n)) + -->
<!--   see::geom_point2(size=2) + -->
<!--   geom_abline(slope=1) + -->
<!--   coord_fixed(ratio = 1, xlim = c(0, 3), ylim = c(0, 3)) + -->
<!--   see::theme_modern() -->
<!-- ``` -->

<!-- It seems like adding the log of the sample size improves the relationship! What is the formula of this new **log-odds** to ***d*** conversion? -->

<!-- ```{r message=FALSE, warning=FALSE} -->
<!-- formula <- insight::get_parameters(model) %>%  -->
<!--   mutate(Coef = paste0(round(Estimate, 5), " * (", Parameter, ")"), -->
<!--          Coef = gsub(":", " * ", Coef), -->
<!--          Coef = gsub(" * ((Intercept))", "", Coef, fixed=TRUE)) %>%  -->
<!--   pull(Coef) %>%  -->
<!--   paste(collapse = " + ") -->

<!-- print(formula) -->
<!-- ``` -->

<!-- ```{r message=FALSE, warning=FALSE} -->
<!-- newconversion <- function(logodds, n){ -->
<!--   0.87341 + 0.75933 * (logodds) + 0.02705 * (n) + -0.35727 * (log(n)) + 0.03714 * (logodds * n) + -0.61004 * (logodds * log(n)) + -0.00418 * (n * log(n)) + -0.00563 * (logodds * n * log(n)) -->
<!-- } -->

<!-- df$d_from_coef_adjusted <- newconversion(df$logodds, df$n) -->

<!-- ggplot(df, aes(x=d_cohen, y=d_from_coef_adjusted)) + -->
<!--   geom_point(size=1) + -->
<!--   geom_abline(slope=1) + -->
<!--   coord_fixed(ratio = 1, xlim = c(0, 3), ylim = c(0, 3)) + -->
<!--   see::theme_modern() -->
<!-- ``` -->


# References