---
title: "Effect Sizes for Simple Hypothesis Tests"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_width: 10.08
    fig_height: 6
tags: [r, effect size, rules of thumb, guidelines, conversion]
vignette: >
  \usepackage[utf8]{inputenc}
  %\VignetteIndexEntry{Effect Sizes for Simple Hypothesis Tests}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
library(knitr)
options(knitr.kable.NA = "")
knitr::opts_chunk$set(comment = ">")
options(digits = 3)

pkgs <- c("effectsize", "BayesFactor")
if (!all(sapply(pkgs, require, quietly = TRUE, character.only = TRUE))) {
  knitr::opts_chunk$set(eval = FALSE)
}
set.seed(7)
```

This vignette provides a short review of effect sizes for common hypothesis
tests (in **`R`** these are usually achieved with various `*.test()`
functions).

```{r}
library(effectsize)
library(BayesFactor)
```

In most cases, the effect sizes can be automagically extracted from the `htest`
object via the `effectsize()` function.

## Standardized Differences

For *t*-tests, it is common to report an effect size representing a standardized
difference between the two compared samples' means. These measures range from
$-\infty$ to $+\infty$, with negative values indicating the second group's mean
is larger (and vice versa).

### Two Independent Samples

For two independent samples, the difference between the means is standardized
based on the pooled standard deviation of both samples (assumed to be equal in
the population):

```{r}
t.test(mpg ~ am, data = mtcars, var.equal = TRUE)

cohens_d(mpg ~ am, data = mtcars)
```

Hedges' *g* provides a bias correction for small sample sizes ($N < 20$).

```{r}
hedges_g(mpg ~ am, data = mtcars)
```

If variances cannot be assumed to be equal, it is possible to get estimates that
are not based on the pooled standard deviation:

```{r}
t.test(mpg ~ am, data = mtcars, var.equal = FALSE)

cohens_d(mpg ~ am, data = mtcars, pooled_sd = FALSE)

hedges_g(mpg ~ am, data = mtcars, pooled_sd = FALSE)
```

In cases where the differences between the variances are substantial, it is also
common to standardize the difference based only on the standard deviation of one
of the groups (usually the "control" group); this effect size is known as Glass'
$\Delta$ (delta) (Note that the standard deviation is taken from the *second* sample).

```{r}
glass_delta(mpg ~ am, data = mtcars)
```

For a one-sided hypothesis, it is also possible to construct one-sided confidence intervals:

```{r}
t.test(mpg ~ am, data = mtcars, var.equal = TRUE, alternative = "less")

cohens_d(mpg ~ am, data = mtcars, pooled_sd = TRUE, alternative = "less")
```

#### Common Language Effect Sizes

Related effect sizes are the *common language effect sizes* which present information about group differences in terms of probability.

```{r}
cles(mpg ~ am, data = mtcars)
```

### One Sample and Paired Samples

In the case of a one-sample test, the effect size represents the standardized
distance of the mean of the sample from the null value. For paired-samples, the
difference between the paired samples is used:

```{r}
t.test(extra ~ group, data = sleep, paired = TRUE)

cohens_d(extra ~ group, data = sleep, paired = TRUE)

hedges_g(extra ~ group, data = sleep, paired = TRUE)
```

### For a Bayesian *t*-test

```{r}
(BFt <- ttestBF(mtcars$mpg[mtcars$am == 0], mtcars$mpg[mtcars$am == 1]))

effectsize(BFt, test = NULL)
```

## One way ANOVA

For more details, see [ANOVA
vignette](https://easystats.github.io/effectsize/articles/anovaES.html).

```{r, message=FALSE}
onew <- oneway.test(mpg ~ gear, data = mtcars, var.equal = TRUE)

eta_squared(onew)
```

## Contingency Tables

### 2-by-2 Tables

For 2-by-2 contingency tables, $\phi$ (Phi) is homologous (though directionless) to the bi-serial correlation between the two dichotomous variables, with 0 representing no association, and 1 representing a perfect association. A "cousin" effect size is Pearson's contingency coefficient.

```{r}
MPG_Gear <- table(mtcars$mpg < 20, mtcars$vs)

phi(MPG_Gear)

# Same as:
cor(mtcars$mpg < 20, mtcars$vs)



pearsons_c(MPG_Gear)
```

(Cramer's *V* or Cohen's *w* can also be used, but for 2-by-2 tables, they are equivalent to $\phi$.)

In addition to $\phi$ and Pearson's *C*, we can also
compute the Odds-ratio (OR), where each column represents a different *group*.
Values larger than 1 indicate that the odds are higher in the first group (and
vice versa).

```{r}
(RCT <- matrix(
  c(
    71, 30,
    50, 100
  ),
  nrow = 2, byrow = TRUE,
  dimnames = list(
    Diagnosis = c("Sick", "Recovered"),
    Group = c("Treatment", "Control")
  )
))

chisq.test(RCT) # or fisher.test(RCT)

oddsratio(RCT)
```

We can also compute the Risk-ratio (RR), which is the ratio between the
proportions of the two groups - a measure which some claim is more intuitive.

```{r}
riskratio(RCT)
```

Additionally, Cohen's *h* can also be computed, which uses the *arcsin*
transformation. Negative values indicate smaller proportion in the first group
(and vice versa).

```{r}
cohens_h(RCT)
```

### Larger Tables

For larger contingency tables Cramér's *V*, Cohen's *w* and Pearson's *C* can be used.
While Cramér's *V* and Pearson's *C* are capped at 1 (perfect association), Cohen's *w* can be larger than 1
(for all three, 0 indicates no association between the variables).

```{r}
(Music <- matrix(
  c(
    150, 130, 35, 55,
    100, 50, 10, 40,
    165, 65, 2, 25
  ),
  byrow = TRUE, nrow = 3,
  dimnames = list(
    Study = c("Psych", "Econ", "Law"),
    Music = c("Pop", "Rock", "Jazz", "Classic")
  )
))

chisq.test(Music)

cramers_v(Music)

cohens_w(Music)

pearsons_c(Music)
```

These can also be extracted from the equivalent Bayesian test:

```{r}
(BFX <- contingencyTableBF(Music, sampleType = "jointMulti"))

effectsize(BFX, type = "cramers_v", test = NULL)

effectsize(BFX, type = "cohens_w", test = NULL)

effectsize(BFX, type = "pearsons_c", test = NULL)
```

### Goodness-of-Fit

Cohen's *w* and Pearson's *C* are also applicable to tests of goodness-of-fit, 
where small values indicate no deviation from the hypothetical probabilities and large values indicate... large deviation from the hypothetical probabilities.

```{r}
O <- c(89, 37, 130, 28, 2) # observed group sizes
E <- c(.40, .20, .20, .15, .05) # expected group freq

chisq.test(O, p = E)

pearsons_c(O, p = E)

cohens_w(O, p = E)
```

However, Cohen's *w* does not account for the distribution of expected probabilities, 
and as such may be seen is an inflated effect size, and since it can be larger than 1 it is also harder to interpret. 
For these reasons, we recommend the $פ$ (Fei) coefficient, which adjusted Cohen's *w* to account for the expected distribution of probabilities, making it range between 0 (observed distribution matches the expected distribution perfectly) and 1 (the observed distribution is maximally different than the expected one).

```{r}
fei(O, p = E)
```

### Paired Contingency Tables

For dependent (paired) contingency tables, Cohen's *g* represents the symmetry
of the table, ranging between 0 (perfect symmetry) and 0.5 (perfect asymmetry).

```{r}
(Performance <- matrix(
  c(
    794, 86,
    150, 570
  ),
  nrow = 2, byrow = TRUE,
  dimnames = list(
    "1st Survey" = c("Approve", "Disapprove"),
    "2nd Survey" = c("Approve", "Disapprove")
  )
))

mcnemar.test(Performance)

cohens_g(Performance)
```

## Rank Based tests

Rank based tests get rank based effect sizes!

### Difference in Ranks

For two independent samples, the rank-biserial correlation ($r_{rb}$) is a
measure of relative superiority - i.e., larger values indicate a higher
probability of a randomly selected observation from *X* being larger than
randomly selected observation from *Y*. A value of $(-1)$ indicates that all
observations in the second group are larger than the first, and a value of
$(+1)$ indicates that all observations in the first group are larger than the
second.

```{r, warning=FALSE}
A <- c(48, 48, 77, 86, 85, 85)
B <- c(14, 34, 34, 77)

wilcox.test(A, B) # aka Mann–Whitney U test

rank_biserial(A, B)
```

Here too we have a *common language effect size*:

```{r}
cles(A, B, rank = TRUE)
```

Or, we can present the difference between ordinal ranks as the odds of scoring
higher in one group than another. This is called a "generalized odds ratio", or 
the Wilcoxon-Mann-Whitney, $WMW_{odds}$, odds (hence the `wmw_odds` function)^[For greater detail, please read the SAS support paper by O'Brien and Castelloe https://support.sas.com/resources/papers/proceedings/proceedings/sugi31/209-31.pdf]. 
Like the  rank-biserial correlation, it is a measure of relative superiority - i.e., 
larger values indicate a higher probability of a randomly selected observation 
from *X* being larger than randomly selected observation from *Y*. A value of 1
indicates complete equivalence between groups (i.e., odds are equal). A value of
$(0)$ indicates that all observations in the second group are larger than the 
first, and a value of $(Inf)$ indicates that all observations in the first group 
are larger than the second.

```{r}
wmw_odds(A, B)
```

For one sample, $r_{rb}$ measures the symmetry around $\mu$ (mu; the null
value), with 0 indicating perfect symmetry, $(-1)$ indicates that all
observations fall below $\mu$, and $(+1)$ indicates that all observations fall
above $\mu$. For paired samples the difference between the paired samples is
used:

```{r}
x <- c(1.15, 0.88, 0.90, 0.74, 1.21, 1.36, 0.89)

wilcox.test(x, mu = 1) # aka Signed-Rank test

rank_biserial(x, mu = 1)


x <- c(1.83, 0.50, 1.62, 2.48, 1.68, 1.88, 1.55, 3.06, 1.30)
y <- c(0.878, 0.647, 0.598, 2.05, 1.06, 1.29, 1.06, 3.14, 1.29)

wilcox.test(x, y, paired = TRUE) # aka Signed-Rank test

rank_biserial(x, y, paired = TRUE)
```

Similarly, $WMW_{odds}$  also measures the symmetry around $\mu$, with 1 
indicating perfect symmetry, $(0)$ indicates that all
observations fall below $\mu$, and $(Inf)$ indicates that all observations fall
above $\mu$.

```{r}
x <- c(1.15, 0.88, 0.90, 0.74, 1.21, 1.36, 0.89)

wmw_odds(x, mu = 1)

x <- c(1.83, 0.50, 1.62, 2.48, 1.68, 1.88, 1.55, 3.06, 1.30)
y <- c(0.878, 0.647, 0.598, 2.05, 1.06, 1.29, 1.06, 3.14, 1.29)

wmw_odds(x, y, paired = TRUE)
```

### Rank One way ANOVA

The Rank-Epsilon-Squared ($\varepsilon^2$) is a measure of association
for the rank based one-way ANOVA. Values range between 0 (no relative
superiority between any of the groups) to 1 (complete separation - with no
overlap in ranks between the groups).

```{r}
group_data <- list(
  g1 = c(2.9, 3.0, 2.5, 2.6, 3.2), # normal subjects
  g2 = c(3.8, 2.7, 4.0, 2.4), # with obstructive airway disease
  g3 = c(2.8, 3.4, 3.7, 2.2, 2.0) # with asbestosis
)

kruskal.test(group_data)

rank_epsilon_squared(group_data)
```

### Rank One way Repeated-Measures ANOVA

For a rank based repeated measures one-way ANOVA, Kendall's *W* is a measure of
agreement on the effect of condition between various "blocks" (the subjects), or
more often conceptualized as a measure of reliability of the rating / scores of
observations (or "groups") between "raters" ("blocks").

```{r}
# Subjects are COLUMNS
(ReactionTimes <- matrix(
  c(
    398, 338, 520,
    325, 388, 555,
    393, 363, 561,
    367, 433, 470,
    286, 492, 536,
    362, 475, 496,
    253, 334, 610
  ),
  nrow = 7, byrow = TRUE,
  dimnames = list(
    paste0("Subject", 1:7),
    c("Congruent", "Neutral", "Incongruent")
  )
))

friedman.test(ReactionTimes)

kendalls_w(ReactionTimes)
```

# References
